---
title: "P8105 Homework 3"
author: "Hyun Kim (hk3373)"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, echo = FALSE, message = FALSE}
library(tidyverse)
library(patchwork)
library(ggridges)

theme_set(theme_minimal() + 
  theme(legend.position = "bottom", 
      plot.title = element_text(hjust = 0.5)))
```

# Problem 1

## Load NY NOAA data:
```{r load_ny_noaa}
library(p8105.datasets)
data("ny_noaa")
str(ny_noaa)
```

The dataset has `r nrow(ny_noaa)` rows and `r ncol(ny_noaa)` columns, including 
id, tmax and tmin as character variables, precipitation, snow as integer 
variables,and date which is a date variable. There is a total of 
`r sum(is.na(ny_noaa))` NA values in the dataset. Therefore, this could be an 
issue in data analysis since there is a significant number of missing data. 

## Tidy NY NOAA: 
```{r tidy_ny_noaa}
ny_noaa = ny_noaa |>
  drop_na() |>
  mutate(
    prcp = prcp / 10,
    tmax = as.integer(tmax) / 10,
    tmin = as.integer(tmin) / 10
  ) |>
  separate(date, 
           into = c("year", "month", "day"), 
           sep = "-",
           remove = FALSE) 
```

In tidying the dataset, the values of precipitation and temperature were divided 
by 10 since they were in tenths of mm and degrees Celsius respectively. 

### For snowfall, what are the most commonly observed values? Why?
```{r most_common_snowfall}
ny_noaa |>
  group_by(snow) |>
  summarize(count = n()) |>
  arrange(desc(count)) |>
  slice(1)
```

For snowfall, the most commonly observed value is 0 because it only snows in 
winter, but the dataset includes data from the other three seasons when it 
most likely doesn't snow. 

## Average max temperature in January and July:
```{r avg_tmax_jan_july} 
ny_noaa |>
  filter(month %in% c("01","07")) |>
  mutate(
    month = recode(month, "01" = "January"),
    month = recode(month, "07" = "July")
  ) |>
  group_by(id, year, month) |>
  summarize(average_tmax = round(mean(tmax), 2)) |>
  ggplot(aes(x = year, y = average_tmax, color = month)) +
  geom_point() +
  facet_grid(~month) +
  theme(axis.text.x = element_text(angle = 90)) +
  ggtitle("Average Max Temperature in January and July")
```

### Is there any observable / interpretable structure? Any outliers?

The average max temperature is noticeably higher in July compared to January 
obviously because it's summer during July and winter during January. Data for 
July 1988 has an extreme outlier. Data for January has noticeable outliers for 
several years; the most extreme outliers are shown in 1982 and 1996. 

## tmax vs tmin and snowfall distribution by year:
```{r tmax_tmin_snowfall, fig.width = 10, fig.asp = .7}
tmax_vs_tmin = 
  ny_noaa |>
  ggplot(aes(x = tmax, y = tmin)) + 
  geom_hex() +
  ggtitle("tmax vs tmin")

snowfall_dist = 
  ny_noaa |>
  filter(snow > 0, snow < 100) |>
  ggplot(aes(x = snow, y = year)) + 
  geom_density_ridges(scale = 1.5) +
  ggtitle("Snowfall Distribution by Year")

tmax_vs_tmin + snowfall_dist
```

According to the figure above, the values of tmax are concentrated around -5 to 
30 degrees Celsius, while the values of tmin are concentrated around -20 to 20 
degrees. The distribution of snowfall for all years are right skewed because it 
only snows during winter and it's subsequently more common for data in the 
dataset to have lower values of snowfall overall. 
